---
title: "Does rurality have an impact of Coronavirus?"
output: html_notebook
---

```{r}
library(tidyverse)

```

Import daily covid cases and deaths for the UK from the Public Health England coronavirus data api.

```{r}

uk_daily_covid19_cases <- read_csv("https://api.coronavirus.data.gov.uk/v2/data?areaType=utla&metric=newCasesByPublishDate&format=csv")

uk_daily_covid19_deaths <- read_csv("https://api.coronavirus.data.gov.uk/v2/data?areaType=utla&metric=newOnsDeathsByRegistrationDate&format=csv")

```

```{r}
glimpse(uk_daily_covid19_cases)

glimpse(uk_daily_covid19_deaths)

```

### Daily Cases

Data has `r nrow(uk_daily_covid19_cases)` rows and `r ncol(uk_daily_covid19_cases)` columns.

Dates: `r min(uk_daily_covid19_cases$date)` to `r max(uk_daily_covid19_cases$date)`.

It assigns covid cases by the date the data was published on, not when the initial test was taken.

The data is at Upper Tier Local Authority level.

### Daily Deaths

Data has `r nrow(uk_daily_covid19_deaths)` rows and `r ncol(uk_daily_covid19_deaths)` columns.

Dates: `r min(uk_daily_covid19_deaths$date)` to `r max(uk_daily_covid19_deaths$date)`.

It assigns deaths to covid if COVID-19 is mentioned as a cause on the death certificate, while the date refers to the date the death was registered.

The data is at Upper Tier Local Authority level.

```{r}



```

